---
title: "French given names per year per department"
author: "Lucas Mello Schnorr, Jean-Marc Vincent"
date: "October, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using the [given names data
set of INSEE](https://www.insee.fr/fr/statistiques/fichier/2540004/dpt2017_txt.zip), answer some of the following questions:

- First name frequency evolves along time?
- What can we say about "the french version of your name here" (for each state, all the country)?
- Is there some sort of geographical correlation with the data?
- Which state has a larger variety of names along time?
- _your own question_ (be creative)

You need to use the _dplyr_ and _magrittr_ for this analysis. Read in R with this code. Note that you might need to install the `readr` package with the appropriate command.

```{r}
file = "dpt2017_txt.zip"
if(!file.exists(file)){
  download.file("https://www.insee.fr/fr/statistiques/fichier/2540004/dpt2017_txt.zip",
	destfile=file)
}
unzip(file)
```

```{r}
library(readr);
df <- read_tsv("dpt2017.txt", locale = locale(encoding = "ISO-8859-1"), col_types=cols(
  sexe = col_integer(),
  preusuel = col_character(),
  annais = col_character(),
  dpt = col_character(),
  nombre = col_integer()
));
df;
```

Finding the occurencies of my name 'Jean' by each year and plotting it to a line graph.

```{r}
  library(tidyverse)
  library(ggplot2)
  df_jean <- df %>% filter(preusuel == 'JEAN') %>% group_by(annais) %>% summarise(quant = sum(nombre))
  
  df_jean <-  df_jean  %>% mutate(annais=as.integer(annais))
  
  head(df_jean)
  
  ggplot(df_jean, aes( x=annais, y =quant))+geom_line()
```

Finding the distribution of sexes.

```{r}
  library(tidyverse)
  sex_df <- df %>% group_by(annais) %>% group_by(sexe) %>% summarise(quant = sum(nombre))
  sex_df
```

Visualizing the distribution of the observations between the dpt's.

```{r}
  library(tidyverse)
  name_df <- df %>% group_by(dpt) %>%  summarise(quant = sum(nombre))
  #name_df
  name_df <- name_df %>% mutate(dpt=as.integer(dpt))
  ggplot(name_df, aes( x=dpt, y =quant))+geom_point() + coord_cartesian(xlim=c(0,100))
```

