---
title: "Handson analysis on the POA accidents dataset"
author: "Joao Pedro Oliveira"
date: "31 de outubro de 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#This is my hands on analysis of the POA accidents dataset.

####First download the dataset
```{r}
file = "acidentes-2016.csv"
if(!file.exists(file)){
  download.file("http://datapoa.com.br/storage/f/2017-08-03T13%3A19%3A45.538Z/acidentes-2016.csv", destfile=file)
}
```

#### Now, read the CSV file to a Dataframe using readr
```{r}
library(readr)
ac_data <- read_delim(file, ";")
ac_data
```

#### We need to get a grasp for what is in our dataset
```{r}
summary(ac_data)
```

####As we see, there is a lot of information here.To work with this data we need to do some cleaning.

Since for this first analysis we'll be trying to find out if there is a time of the year with more accidents, we'll limit this dataset for this pourpose.

Here we group our data by date and summarise it to a dataframe with only two columns. Then, we plot it to see the number of occurences by date.

```{r}
ac_data %>%  
  group_by(DATA) %>% 
  summarise(QUANT_ACID = sum(CONT_ACID)) %>%
  ggplot(aes( x=DATA, y =QUANT_ACID))+geom_line() + geom_point()
```
```{r}
ac_data %>%  
  group_by(DATA) %>% 
  summarise(QUANT_ACID = sum(CONT_ACID)) %>%
  ggplot(aes( x=DATA, y =QUANT_ACID))+geom_col()
```
####As we can see, there seems to be no correlation between the time of the year and the accidents.

```{r}
ac_data %>%  
  group_by(MES) %>% 
  summarise(QUANT_ACID = sum(CONT_ACID)) %>%
  ggplot(aes( x=MES, y =QUANT_ACID))+geom_col(count=12,binwidth = 1)
```
#### Now, let's analyse to learn how many vehicles are usually involved in the accidents.

```{r}
ac_data %>%  
  group_by(AUTO) %>% 
  summarise(QUANT_ACID = sum(CONT_ACID)) %>%
  ggplot(aes( x=AUTO, y =QUANT_ACID))+geom_col()
```
####So we can see from here that most accidents happen envolving 1 or 2 vehicles.

####Now let's see if there is a certain weekday that has more accidents than others

```{r}
ac_data %>%  
  group_by(DIA_SEM) %>% 
  summarise(QUANT_ACID = sum(CONT_ACID)) %>%
  ggplot(aes( x=DIA_SEM, y=QUANT_ACID))+geom_col()
```

#### From this graph we can certainly observe some interesting things. The first thing that comes to mind is that there are more accidents on Fridays, usually when people go out to party. And the number of accidents on Saturdays and Sundays are low, maybe because people tend to stay at home during those days.

#### So, lets see if there are regions in Porto Alegre with more accidents than others. For this, I define "Region" as the column 'REGIAO'.

```{r}
ac_data %>%  
  group_by(REGIAO) %>% 
  summarise(QUANT_ACID = sum(CONT_ACID)) %>%
  ggplot(aes( x=REGIAO, y=QUANT_ACID))+geom_col()
```

####Now, this seems to be good enough. One thing i'm concerned, though,it's to see the percentages of the total accidents by region. This might be more interesting.

```{r}
ac_data %>%  
  group_by(REGIAO) %>% 
  summarise(PERCNT_ACID = sum(CONT_ACID)/nrow(ac_data) * 100) %>% 
  ggplot(aes( x=REGIAO, y=PERCNT_ACID))+geom_col()
```
